The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-11 kalman_walk_paper.tex
edited examples

\newcommand{\Sys}{\mathcal{S}}

@@

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

@@

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
  \begin{align}
    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

Two systems, even if in different dimensions, will have identical phenotypes if they are in $\mathcal{K}$. Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is \emph{all} gene regulatory networks can drift/rewire. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.

A large, adapted population will move around in $\mathcal{K}$ given environmental and selective stasis. Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
    %\begin{align*}
    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
    %\end{align*}
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ ($A(0)$) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ ($A(2)$).}
  \end{figure}
\end{example}

@@

For instance, diploid system coefficients are the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} = \left(\Sys + \Sys'\right)/2$. Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(\Sys \times \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.

Even if two genomes $\Sys$ and $\Sys'$ are both in $\mathcal{K}$, their diploid and recombinant systems may not be in $\mathcal{K}$. This is because $\mathcal{K}$ is not required to be closed under averaging or recombination.

If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes. 
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
Each $A$ is phenotypically identical ($\phi(t) = \sin(t) + \cos(t)$), however some of the hybrids exhibit markedly different dynamics.

@@

>>>>

\newcommand{\Sys}{\mathcal{S}}
\newcommand{\allS}{\mathcal{K}(h)}

@@

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response,
      \begin{align*}
        \phi(t) = h(t) = \sin t + \cos t  .
      \end{align*}

@@

More generally,
we denote by $\calA_n(\Sys_0)$ the set of all $n$-dimensional systems equivalent to $\Sys_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\Sys_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.
Further, we refer to the set of all linear systems with impulse response $h(t)$, regardless of dimension, to be $\allS$,
  \begin{align}
    \allS := \bigcup_{n=\min}^{\infty} \calA_n(\Sys_0)
  \end{align}
Two systems, even if in different dimensions, will have identical phenotypes if they are in $\allS$. Further, there is no unique triple $(A,B,C)$ for the mapping $h: u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is under these assumptions, \emph{all} gene regulatory networks can drift/rewire. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition}\ref{apx:kalman}.

A large, adapted population will move around in $\allS$ under environmental and selective stasis. In the minimal dimension neutral evolution proceeds as a stochastic coordinate transformation on the elements in $\Sys_{\min}$. Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.

@@

%\plr{This is not part of the definition of $\mathcal{I}$:}
%\js{should we delete this example ($A(0) \times A(2)$ hybrids)? I like how clean it is -- but I'm not sure it's that important in this context. It's a good proof of principle though.}
%\jss{$F_1$s created by crossing phenotypically equivalent oscillators $A(0)$ and $A(2)$ have a phenotype of $\phi_{F_{1}}(t) = e^{t}$, 
%in contrast to both parents, who have $\phi_{P_{1}}(t) = \phi_{P_{2}}(t) = \sin(t) + \cos(t)$. 
%The hybrid phenotype is significantly different (it does not oscillate and increases infinitely) despite the phenotypic equivalence of the parents.
%%\plr{don't need these to be big equations.}
%%\begin{align*}
%%  \fit \left(\phi_{P_{1}}\right) = \fit \left( \phi_{P_{2}} \right) &= 1 \\
%%  \fit\left(\phi_{F_{1}}\right) &= 0 \\
%%  \mathcal{I} &= 0
%%\end{align*}
%
%%Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
%
%  \begin{figure}[H]
%    \centering
%    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
%    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
%  \end{figure}
%  }
%%     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
%%       DMI examples...
%%       \begin{figure}[H]
%%         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
%%       \end{figure}
%%     \end{example}
%
%%     \begin{example}[Not all Networks can Host Incompatibilities]
%%       convex sets cant have DMIs
%
%%       \begin{align*}
%%         h(t) = 2 e^{- \theta t}
%%       \end{align*}
%%       Any non-minimal system with rows summing to $\theta$ is PI. Further, these systems are closed under averaging (mating) and row swapping (meiosis), leaving all hybrids optimally fit. The set of gene matrices is affine and therefore convex.  
%%     \end{example}
%
%
%\plri{Missing: a statement about when there are neutral directions, and how many.}
%
%\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}
%
%
 \begin{example}[All Phenotypically Equivalent Cell Cycle Control Networks] \label{ex:all_osc}
    The set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in \ref{ex:oscillator}, where only $A$ can vary, is given by
    %\begin{align*}
    %  A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
    %\end{align*}
    \begin{align*}
      A(\tau) &= \frac{1}{\tau-1} \begin{bmatrix} \tau & -1 \\ 2 \tau(\tau - 1) + 1 &  -\tau \end{bmatrix} \ \text{for} \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

 % \begin{example}[External Equivalence does not imply internal equivalence]
   Despite the phenotypic equivalence of all instantiations of $A(\tau)$, the kryptotypes, vary as a function of $\tau$. 
    Gene-1 dynamics (blue) are equivalent for network architectures $A(0)$ and $A(2)$, however the dynamics of gene-2 (orange) differ with $\tau$.
  \begin{figure}[H]
    \centering
   % \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{osc_A0_A2_both_compare}
    %\caption{Gene-1 (blue) and gene-2 (orange) dynamics for $A(0)$ (top) and $A(2)$ (bottom). Both (top and bottom) gene-1 dynamics are given by $ \kappa_{1} = \sin(t) + \cos(t)$, and gene-2 by $\kappa_{2} = \cos(t) - \sin(t)$ (top) and $\kappa'_{2} = \cos(t) + 3 \sin(t)$ (bottom).}
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{figures/osc_kryp_compare}
    \caption{(Gene-1 dynamics (blue) for both systems $A(0)$ and $A(2)$ are identical, however, $A(0)$ gene-2 dynamics (orange) differ from $A(2)$ (green). Both gene-1 dynamics are given by $\kappa_{1} = \sin t + \cos t$, and gene-2 by $\kappa_{2} = \cos t - \sin t$ ($A(0)$) and $\kappa'_{2} = \cos t + 3 \sin t$ ($A(2)$).}
  \end{figure}
\end{example}

@@

For instance, diploid system coefficients are derived by taking the average of two haploid genomes $\Sys$ and $\Sys'$: $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$. During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.

Even if two genomes ($\Sys$ and $\Sys'$) have the same phenotype, their diploid and recombinant systems may not. This is because $\allS$ is not required to be closed under averaging or recombination.

If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.

\paragraph*{Hybrid phenotypic breakdown between oscillators}
For example, if the two phenotypically equivalent systems $A(0)$ and $A(2)$ from Example \ref{ex:all_osc} hybridize, their offspring's phenotype will fail to oscillate completely. 
  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{expF1}
    \caption{$F_1$ hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid $\left(A_{F_1}\!\!=\!\!\tfrac{A\!(\!0\!) \! + \! A\!(\!2\!)}{2}\right)$ fails to oscillate and exhibits qualitatively different dynamics ($\phi_{F_0}\! = \! \sin t \! + \! \cos t$, however $\phi_{F_1} \! = \! \exp t$).
    }
  \end{figure}
%
%\paragraph{Sexual reproduction and speciation}
%A diploid organism contains two gene network copies: one from each parent. As such, a diploid's system dynamics is computed by averaging the coefficients of both systems. To reproduce, each organism passes on a recombined haploid gene network to its offspring. Haploid gene networks swap system matrix rows randomly between its own two networks.
%Therefore the phenotypic dynamics typical of a first generation ($F_{1}$) cross between two allopatric populations will be determined simply by averaging it's two parental genomes. However, in second generation hybrid ($F_2$) crosses, first new haploid systems will be formed by recombination -- in the process shuffling and combining regulatory coefficients from allopatric populations -- and then brought together to form a diploid.
%
%
  \begin{example}[Hybrid Incompatibility in an Oscillating Gene Network] \label{ex:hybrid_osc}

Here we compare the phenotypes for F2 hybrids formed by crossing oscillators $A(2)$ with $A(2.01)$, $A(2.1)$, and $A(2.5)$ ($B$ and $C$ are the same as above). 
Each $A$ is phenotypically identical ($\phi(t) = \sin t + \cos t$), however some of the hybrids exhibit markedly different dynamics.

@@

\section{Sexual reproduction in linear systems}\label{apx:reproduction}

  Recombination: gamete element $A''$ is formed by recombining (randomly swapping rows) between two parental, possibly distinct, system elements $A$ and $A'$ as follows,
  \begin{align*}
    A'' = MA + (I-M)A'
  \end{align*}
  where $M$ is a diagonal matrix with only $0$s and $1$s on the diagonal.
